Taking Dyads Seriously

### 2018-02-16 paper/1_intro.tex
intro and model

The framework that we present advances statistical analysis of dyadic data by accounting for observational dependence while allowing scholars to test the substantive effect of variables of interest. Thus, the AME approach can be used by scholars in the field to continue to generate substantive insights, while having the benefit of being able to better account for the data generating process behind political events of interest. Most importantly, the AME approach concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.

>>>>

The framework that we present advances statistical analysis of dyadic data by accounting for observational dependence while allowing scholars to test the substantive effect of variables of interest. Thus, the AME approach can be used by scholars in the field to continue to generate substantive insights, while being able to account for the data generating process behind political events of interest. Most importantly, the AME approach concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.

### 2018-02-16 paper/2_model.tex
intro and model

Scholars working with dyadic data typically begin by stacking observations associated with each dyad on top of one another. This makes sense if each observation is independent of the others. For example, a conflict initiated from the United States against Japan is assumed to be independent of any conflictual action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While most scholars begin with the assumption that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false both in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause.

@@

Using the structure of an adjacency matrix, Figure~\ref{fig:adjMatDeps} visualizes the types of first- and second-order dependencies that can complicate the analysis of relational data in traditional GLMs. The adjacency matrix on the top left highlights a particular row to illustrate that these values may be more similar to each other than other values because each has a common sender. Interactions involving a common sender also manifest heterogeneity in how active actors are across the network when compared to each other. In most relational datasets (e.g., trade flows, conflict, participation in international organizations, even networks derived from Twitter or Facebook), we often find that there are some actors that are much more active than others \citep{barabasi:reka:1999}. For example, in an analysis of international trade certain countries (e.g., China) export in much larger volumes than other countries for a variety of structural, contextual, and idiosyncratic reasons. Unless one is able to develop a model that can account for the variety of explanations that may play a role in determining why a particular actor is more active than others, parameter estimates from standard statistical models will be biased.\footnote{In an undirected setting instead of studying sender and receiver heterogeneity, we would just be concerned with actor heterogeneity in general.}

\begin{figure}
	\begin{tabular}{cc}
		\textsc{Sender heterogeneity} & \textsc{Receiver Heterogeneity} \\
		\includegraphics[width=.4\textwidth]{adjRowDep.png} & \includegraphics[width=.4\textwidth]{adjColDep.png} \\
		\textsc{Sender-Receiver Covariance} & \textsc{Reciprocity} \\
		\includegraphics[width=.4\textwidth]{adjRowColCovar.png} & \includegraphics[width=.4\textwidth]{adjRecip.png} \\
	\end{tabular}
	\caption{Nodal and dyadic dependencies in relational data.}
	\label{fig:adjMatDeps}
\end{figure}

For similar reasons one also needs to take into account the shared dependence between observations that share a common receiver. The bottom-left panel in Figure~\ref{fig:adjMatDeps} illustrates that sender and receiver type dependencies can also blend together. Specifically, actors who are more likely to send ties in a network tend to also be more likely to receive them. As a result, the rows and columns in an adjacency matrix are often correlated. For example, consider trade flows both from and to many wealthy, developed countries. The bottom-right panel highlights a second-order dependence, specifically, reciprocity. This is a dependency occurring within dyads involving the same actors whereby values of $y_{ij}$ and $y_{ji}$ are correlated. The concept of reciprocity has deep roots in the study of relations between states \citep{richardson:1960,keohane:1989}.

For most relational data, however, dependencies do not simply manifest at the nodal or dyadic level. More often we find significant evidence of higher-order structures that result from dependencies between multiple groups of actors. These dependencies arise because there may be some set of latent attributes between actors that affects their probability of interacting with one another \citep{zinnes:1967,wasserman:faust:1994}. In Figure~\ref{fig:thirdDeps} we provide a visualization of a simulated relational dataset wherein the nodes designate actors and edges between the nodes indicate that an interaction between the two took place. To highlight third-order dependence patterns, nodes with similar latent attributes are colored similarly.

\begin{figure}[ht]
	\includegraphics[width=.6\textwidth]{stochEquiv_v2.pdf}
	\caption{Visualization of a network with higher-order dependence patterns.}
	% need a simpler caption here
	\label{fig:thirdDeps}
\end{figure}

The visualization illustrates that the actors belonging to the same group have a higher likelihood of interacting with each other, whereas interactions across groups are rarer. A prominent example of a network with this type of structure is discussed by \citet{adamic:glance:2005}, who visualize linkages between right and left leaning political blogs preceding the 2004 United States Presidential Election. \citeauthor{adamic:glance:2005} find that the degree of interaction between right and left leaning blogs was minimal, and that most blogs were linked to others that were politically similar. This showcases the types of higher-order dependencies that can emerge in relational data. First, the fact that interactions were determined by a shared attribute, in this case political ideology, is an example of \textit{homophily}. Homophily explains the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which also have a long history in international relations. The other major type of meso-scopic features that emerge in relational data is community structure \citep{mucha:etal:2010}, which is often formalized through the concept of stochastic equivalence \citep{anderson:etal:1992}. Stochastic equivalence refers to a type of pattern in which actors can be divided into groups such that members of the same group have similar patterns of relationships. In the example we have laid out above, each of the left leaning blogs would be considered stochastically equivalent to one another because any given left-leaning blog is more likely to interact with a blog of a similar political position and less likely to interact with one with a divergent political position.

These types of patterns frequently emerge in IR contexts as well.\footnote{For example, see: \citet{manger:etal:2012, kinne:2013, chyzh:2016}.} For example, a perennial finding in the interstate trade literature emphasizes the role that geography plays in determining trade flows. Geographic proximity in the network context is an example of homophily --- a shared attribute between actors that corresponds to a greater likelihood of the event of interest taking place. Alternatively, in the interstate conflict literature, we may find that actors who are each a member of a particular (formal or informal) alliance are likely to act similarly in the conflict network. Specifically, they will tend to initiate conflictual events with actors that their fellow alliance members initiate conflict with, and they will be unlikely to initiate conflict with members of their alliance --- an example of stochastic equivalence. In both these examples, we are able to explicitly parameterize the attribute that might explain the emergence of higher order dependence patterns, but it is likely that at times we will not able to account for the variety of reasons why higher order dependence patterns in networks may develop.

@@

where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver). We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, using first a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$). Next, to account for the dependencies that emerge in dyadic data and that may complicate inference on the parameter associated with exogenous covariates, we add parameters from the SRRM and LFM. $a_{i}$ and $b_{j}$ in Equation~\ref{eqn:ame} represent sender and receiver random effects incorporated from the SRRM framework:

@@

The sender and receiver random effects are modeled jointly from a multivariate normal distribution to account for correlation in how active an actor is in sending and receiving ties. Heterogeneity in the the sender and receiver effects is captured by $\sigma_{a}^{2}$ and $\sigma_{b}^{2}$, respectively, and $\sigma_{ab}$ describes the linear relationship between these two effects (i.e., whether actors who send [receive] a lot of ties also receive [send] a lot of ties). Beyond these first-order dependencies, second-order dependencies are described by $\sigma_{\epsilon}^{2}$ and a within dyad correlation, or reciprocity, parameter $\rho$.

The LFM contribution to the AME is in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. The construction of the LFM here is actually quite similar to the recommender systems that companies like Amazon and Netflix have used to model customer behavior \citep{resnick:varian:1997,bennett:etal:2007}. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When computing the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both positive and negative homophily.}

@@

The presence of these dependencies in relational data points to the fact that there is a complex structure underlying the dyadic events that we observe, and that accounting for this structure is necessary if we are to adequately represent the data generating process. Of course, if one can specify each of the nodal, dyadic, and triadic set of attributes that influence interactions then the conditional independence assumption underlying standard approaches will be satisfied. However, it is rarely the case that this is possible even for TSCS data and thus modeling decisions must account for underlying structure. Failing to account for the underlying structure in either TSCS or dyadic data leads to a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. Additionally, by ignoring these potential interdependencies, we often ignore substantively interesting features of the problem. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?

>>>>

Scholars working with dyadic data typically begin by stacking observations associated with each dyad on top of one another. This makes sense if each observation is independent of the others. For example, a conflict initiated from the United States against Japan is assumed to be independent of any conflictual action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While most scholars begin with the assumption that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false both in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause.

@@

Using the structure of an adjacency matrix, Figure~\ref{fig:adjMatDeps} visualizes the types of first- and second-order dependencies that can complicate the analysis of relational data in traditional GLMs. The adjacency matrix on the top left highlights a particular row to illustrate that these values may be more similar to each other than other values because each has a common sender. Interactions involving a common sender also manifest heterogeneity in how active actors are across the network when compared to each other. In most relational datasets (e.g., trade flows, conflict, participation in international organizations, even networks derived from Twitter or Facebook), we often find that there are some actors that are much more active than others \citep{barabasi:reka:1999}. For example, in an analysis of international trade certain countries (e.g., China) export much larger volumes than other countries for a variety of structural, contextual, and idiosyncratic reasons. Unless one is able to develop a model that can account for the variety of explanations that may play a role in determining why a particular actor is more active than others, parameter estimates from standard statistical models will be biased.\footnote{In an undirected setting instead of studying sender and receiver heterogeneity, we would just be concerned with actor heterogeneity in general.}

\begin{figure}
	\begin{tabular}{cc}
		\textsc{Sender Heterogeneity} & \textsc{Receiver Heterogeneity} \\
		\includegraphics[width=.4\textwidth]{adjRowDep.png} & \includegraphics[width=.4\textwidth]{adjColDep.png} \\
		\textsc{Sender-Receiver Covariance} & \textsc{Reciprocity} \\
		\includegraphics[width=.4\textwidth]{adjRowColCovar.png} & \includegraphics[width=.4\textwidth]{adjRecip.png} \\
	\end{tabular}
	\caption{Nodal and dyadic dependencies in relational data.}
	\label{fig:adjMatDeps}
\end{figure}

For similar reasons one also needs to take into account the dependence between observations that share a common receiver. The bottom-left panel in Figure~\ref{fig:adjMatDeps} illustrates that sender and receiver type dependencies can also blend together. Specifically, actors who are more likely to send ties in a network tend to also be more likely to receive them. As a result, the rows and columns in an adjacency matrix are often correlated. For example, consider trade flows both from and to many wealthy, developed countries. The bottom-right panel highlights a second-order dependence, specifically, reciprocity. This is a dependency occurring within dyads involving the same actors whereby values of $y_{ij}$ and $y_{ji}$ are correlated. The concept of reciprocity has deep roots in the study of relations between states \citep{richardson:1960,keohane:1989}.

For most relational data, however, dependencies do not simply manifest at the nodal or dyadic level. More often we find significant evidence of higher-order structures that result from dependencies between multiple groups of actors. These dependencies arise because there may be a set of latent attributes between actors that affects their probability of interacting with one another \citep{zinnes:1967,wasserman:faust:1994}. In Figure~\ref{fig:thirdDeps} we provide a visualization of a simulated relational dataset wherein the nodes designate actors and edges between the nodes indicate that an interaction between the two took place. To highlight third-order dependence patterns, nodes with similar latent attributes are colored similarly.

\begin{figure}[ht]
	\includegraphics[width=.6\textwidth]{stochEquiv_v2.pdf}
	\caption{Higher-order dependence patterns in a network.}
	% need a simpler caption here
	\label{fig:thirdDeps}
\end{figure}

The visualization illustrates that the actors belonging to the same group have a higher likelihood of interacting with each other, whereas interactions across groups are rarer. A prominent example of a network with this type of structure is discussed by \citet{adamic:glance:2005}, who visualize linkages between right and left leaning political blogs preceding the 2004 United States Presidential Election. \citeauthor{adamic:glance:2005} find that the degree of interaction between right and left leaning blogs is minimal and that most blogs are linked to others that are politically similar. This showcases the types of higher-order dependencies that can emerge in relational data. First, the fact that interactions are determined by a shared attribute, in this case political ideology, is an example of \textit{homophily}. Homophily explains the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which also have a long history in international relations. The other major type of meso-scopic features that emerge in relational data is community structure \citep{mucha:etal:2010}, which is often formalized through the concept of stochastic equivalence \citep{anderson:etal:1992}. Stochastic equivalence refers to a type of pattern in which actors can be divided into groups such that members of the same group have similar patterns of relationships. In the example above, each of the left leaning blogs would be considered stochastically equivalent to one another because any given left-leaning blog is more likely to interact with a blog of a similar political position and less likely to interact with one with a divergent political position.

These types of patterns frequently emerge in IR contexts as well.\footnote{For example, see: \citet{manger:etal:2012, kinne:2013, chyzh:2016}.} For example, a perennial finding in the interstate trade literature emphasizes the role that geography plays in determining trade flows. Geographic proximity in the network context is an example of homophily --- a shared attribute between actors that corresponds to a greater likelihood of the event of interest taking place. Alternatively, in the interstate conflict literature, we may find that actors who are each a member of a particular (formal or informal) alliance are likely to act similarly in the conflict network. Specifically, they will tend to initiate conflictual events with actors that their fellow alliance members initiate conflict with, and they will be unlikely to initiate conflict with members of their alliance --- an example of stochastic equivalence. In both these examples, we are able to explicitly parameterize the attribute that might explain the emergence of higher order dependence patterns. While sometimes the conditions driving these patterns, such as geography, are easy to identify, at other times it can be difficult to describe exactly why higher order dependence patterns in networks may develop.
%CD to SM ^

@@

where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver) at time $t$. We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, first using a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$). Next, to account for the dependencies that emerge in dyadic data that may complicate inference on the parameter associated with exogenous covariates, we add parameters from the SRRM and LFM. $a_{i}$ and $b_{j}$ in Equation~\ref{eqn:ame} represent sender and receiver random effects incorporated from the SRRM framework:

@@

The sender and receiver random effects are modeled jointly from a multivariate normal distribution to account for correlation in how active an actor is in sending and receiving ties. Heterogeneity in the sender and receiver effects is captured by $\sigma_{a}^{2}$ and $\sigma_{b}^{2}$, respectively, and $\sigma_{ab}$ describes the linear relationship between these two effects (i.e., whether actors who send [receive] a lot of ties also receive [send] a lot of ties). Beyond these first-order dependencies, second-order dependencies are described by $\sigma_{\epsilon}^{2}$ and a within dyad correlation, or reciprocity, parameter $\rho$.

The LFM contribution to the AME is in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. The construction of the LFM here is actually quite similar to the recommender systems that companies like Amazon and Netflix use to model customer behavior \citep{resnick:varian:1997,bennett:etal:2007}. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When computing the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model the singular values are not restricted to be positive, thus allowing us to account for both positive and negative homophily.}

@@

The presence of these dependencies in relational data points to the fact that there is a complex structure underlying the dyadic events that we observe, and that accounting for this structure is necessary if we are to adequately represent the data generating process. Of course, if one can specify each of the nodal, dyadic, and triadic set of attributes that influence interactions then the conditional independence assumption underlying standard approaches will be satisfied. However, it is rarely the case that this is possible even for TSCS data and thus modeling decisions must account for underlying structure. Failing to do so in either TSCS or dyadic data leads to a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. Additionally, by ignoring these potential interdependencies, we often ignore substantively interesting features of the phenomena we investigate. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?
